---
title: "Education Data Pre-Processing + Exploration"
author: "Pooja Desai (pmd2137)"
date: "12/10/2022"
output: github_document
---

```{r import, echo=FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(purrr)
```

```{r education_data_preprocessing, paged.print=TRUE}
data <- read.csv("data/census_data/edu_attainment.csv", skip = 1) %>%
  janitor::clean_names() %>%
  mutate(
    edu_pop_total = total_pop_18_24 + total_pop_25 + total_pop_65,
    edu_pop_me = me_pop_18_24 + me_total_pop_25 + me_pop_65, 
    edu_pop_under65_total = total_pop_18_24 + (total_pop_25 - total_pop_65),
    edu_pop_under65_me = me_pop_18_24 + (me_total_pop_25 - me_pop_65),
    
    #regroup 18-24 into edu categories
    total_hs.under_18_24 = total_less_hs12_18_24 + total_hs_18_24,
    total_hs.to.ba_18_24 = total_some_col_18_24,
    total_ba.over_18_24 = total_col_18_24,

    #regroup 25+ into edu categories
    total_hs.under_25.over = total_pop_25 - total_over_hs_25,
    total_hs.to.ba_25.over = total_over_hs_25 - total_ba_over_25,
    total_ba.over_25.over = total_ba_over_25,

    #regroup 65+ into edu categories
    hs.under_65.over = total_pop_65 - total_over_hs_65,
    hs.to.ba_65.over= total_over_hs_65 - total_ba_over_65,
    ba.over_65.over = total_ba_over_65,
    
    #calculate under 65, into edu categories
    hs.under_under.65 = total_hs.under_18_24 + (total_hs.under_25.over - hs.under_65.over),
    hs.to.ba_under.65 = total_hs.to.ba_18_24 + (total_hs.to.ba_25.over - hs.to.ba_65.over),
    ba.over_under.65 = total_ba.over_18_24 + (total_ba.over_25.over - ba.over_65.over),
    modzcta = substr(zcta_zip,6,11)) %>% #extract ZCTA zipcode
  select(
    modzcta, edu_pop_total, edu_pop_under65_total, total_pop_65,
    hs.under_under.65, hs.to.ba_under.65, ba.over_under.65,
    hs.under_65.over, hs.to.ba_65.over, ba.over_65.over) -> dataset_clean #export truncated dataset

```

Population buckets: total population (18+) // under 65 population (18-65) // over 65 population (65+)

Educational bucket:

| EDU // AGE bucket   | 18-24            | 25+            | 65+                                 |
|-----------------|--------------------|------------------|------------------|
|--------------------|--------------------|----------------|-----------------|

| HS (GED) or less    | total_less_hs_12_18_24 + total_hs_18_24 | total_pop_25 - total_over_hs_25  | total_pop_65 - total_over_hs_65     |
| btwn HS and college | total_some_col_18_24                    | total_over_hs_25 - total_ba_over_25 | total_over_hs_65 - total_over_ba_65 |
| 4 yr degree +       | total_col_18_24                         | total_over_ba_25                    | total_over_ba_65                    |

```{r vaccine-data}

#read in booster data
vaccine_data <- read.csv("data/coverage-by-modzcta-adults_03.09.22.csv")
zipcode_data <- read.csv("data/Modified_Zip_Code_Tabulation_Areas__MODZCTA_.csv")

#clean vaccine data
vaccine_data <- read.csv("data/coverage-by-modzcta-adults_03.09.22.csv") %>%
  janitor::clean_names() %>%
  mutate(modzcta = as.character(modzcta))

vaccine_data$modzcta <- gsub('\\s+', '', vaccine_data$modzcta)
dataset_clean$modzcta <- gsub('\\s+', '', dataset_clean$modzcta)
```


```{r merge-dataset}
#merge booster and edu
joined_dataset_education = merge(vaccine_data,dataset_clean, by = "modzcta", all.x = TRUE)

joined_dataset_education %>%
  mutate(
    hs.under_total = hs.under_under.65 + hs.under_65.over,
    hs.to.ba_total = hs.to.ba_under.65 + hs.to.ba_65.over,
    ba.over_total = ba.over_under.65 + ba.over_65.over) %>%
  
  pivot_longer(c("hs.under_under.65", "hs.to.ba_under.65","ba.over_under.65", 
                 "hs.under_65.over", "hs.to.ba_65.over","ba.over_65.over",
                 "hs.under_total","hs.to.ba_total", "ba.over_total"), 
               names_to = "age_edu_label", 
               values_to = "edu_count") %>%
  
  separate(col=age_edu_label, into=c("edu","age"), sep="_", convert = TRUE) %>%
  mutate(edu = recode(edu, "hs.under" = "HS or less", 
                      "hs.to.ba" = "HS to B.A.", "ba.over" = "B.A. or more"),
    edu = factor(edu),
    age = recode(age, "under.65" = "under 65", "65.over" = "65 +"),
    age = factor(age),
    borough = factor(borough),
    count_age_edu = ifelse(age == "total", edu_pop_total, 
                       ifelse(age == "under 65",edu_pop_under65_total,total_pop_65)),
    edu_prop = edu_count/count_age_edu,
    partially_vaccinated_prop = count_partially_cumulative/pop_denominator,
    fully_vaccinated_prop = count_fully_cumulative/pop_denominator,
    booster_prop = count_additional_cumulative/pop_denominator) %>%
  
    #select variables for final merged dataset
    select(borough, modzcta, neighborhood_name, pop_denominator, count_age_edu,
           edu_count, edu, age, partially_vaccinated_prop,
           fully_vaccinated_prop, booster_prop, edu_prop) -> edu_booster_dataset
```


```{r export_csv}
write.csv(edu_booster_dataset, "./data/merged/cleaned_edu_booster_dataset.csv", row.names=FALSE)
```

### Fully Vaccinated Individuals by Education Status and Age
``` {r, echo = FALSE}
ggplot(edu_booster_dataset) +
  aes(x = edu_prop, y = fully_vaccinated_prop, colour = edu) +
  geom_point(shape = "circle", size = 1.5) +
  scale_color_hue(direction = 1) +
  labs(
    x = "Proportion of Individuals by Education Status",
    y = "Proportion of Fully Vaccinated Individuals ",
    title = "Fully Vaccinated Individuals by Education Status and Age",
    color = "Education Status"
  ) +
  theme_minimal() +
  facet_wrap(vars(age))
```

### Additional/Booster Individuals by Education Status and Age
``` {r, echo=FALSE}
ggplot(edu_booster_dataset) +
  aes(x = edu_prop, y = booster_prop, colour = edu) +
  geom_point(shape = "circle", size = 1.5) +
  scale_color_hue(direction = 1) +
  labs(
    x = "Proportion of Individuals by Education Status",
    y = "Proportion of Individuals Boosted ",
    title = "Booster Individuals by Education Status and Age",
    color = "Education Status"
  ) +
  theme_minimal() +
  facet_wrap(vars(age))
```

### Fully Vaccinated Individuals by Education Status and Boroughs
```{r, echo = FALSE}
ggplot(edu_booster_dataset) +
  aes(x = edu_prop, y = fully_vaccinated_prop, colour = borough) +
  geom_point(shape = "circle", size = 1.5) +
  scale_color_hue(direction = 1) +
  labs(
    x = "Education Status",
    y = "Proportion of individuals fully vaccinated",
    title = "Fully vaccinated individuals by Education Status and Borough",
    color = "Boroughs"
  ) +
  theme_minimal() +
  facet_wrap(vars(edu))
```

### Additional/Booster Individuals by Education Status and Boroughs
```{r, echo=FALSE}
ggplot(edu_booster_dataset) +
  aes(x = edu_prop, y = booster_prop, colour = borough) +
  geom_point(shape = "circle", size = 1.5) +
  scale_color_hue(direction = 1) +
  labs(
    x = "Education Status",
    y = "Proportion of individuals receiving boosters",
    title = "Additional/Booster individuals by Education Status and Borough",
    color = "Boroughs"
  ) +
  theme_minimal() +
  facet_wrap(vars(edu))

```




